\documentclass[a4paper,12pt,oneside]{book}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
%\usepackage[T1]{fontenc}


% invisible hyperlinks:
\usepackage{hyperref}  

\hypersetup{
  pdftitle={OpenRocket technical documentation},
  pdfauthor={Sampo Niskanen},
  pdfsubject={Technical documentation of the OpenRocket simulation software},
  pdfkeywords={OpenRocket, model rocket, rocketry, simulation, technical documentation},
  pdfpagemode=UseNone,
  colorlinks,
  linkcolor=black,
  filecolor=black,
  urlcolor=black,
  citecolor=black,
  breaklinks=true
}

\usepackage{breakurl}

\usepackage{epsfig}
\usepackage{commath}
\usepackage{ar}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{textcomp}
\usepackage{rotating}
\usepackage{setspace}

\usepackage{array}





\setlength{\parindent}{0mm}
\setlength{\parskip}{\baselineskip}

\newcommand{\ie}{{\it i.e.}\ }
\newcommand{\eg}{{\it e.g.}\ }

\newcommand{\half}{\ensuremath{^1\!/\!_2}}
\newcommand{\quarter}{\ensuremath{^1\!/\!_4}}

\newcommand{\CNa}{\ensuremath{{C_{N_\alpha}}}}
\newcommand{\CNap}{\ensuremath{{C_{N_{\alpha'}}}}}
\newcommand{\Cma}{\ensuremath{C_{m_\alpha}}}
\newcommand{\Aref}{\ensuremath{A_{\rm ref}}}
\newcommand{\Afin}{\ensuremath{A_{\rm fin}}}
\newcommand{\Abase}{\ensuremath{A_{\rm base}}}
\newcommand{\um}{\textmu m}

\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\vi}{\mathbf{i}}
\newcommand{\vj}{\mathbf{j}}
\newcommand{\vk}{\mathbf{k}}

% A space suitable delimiting numbers as  100\s000 for '100 000'
\newcommand{\s}{\nolinebreak\hspace{0.5mm}\nolinebreak}  

\newlength{\numwidth}
\settowidth{\numwidth}{0}
\newcommand{\num}{\hspace{\numwidth}}

\newcommand{\code}[1]{{\tt #1}}





%\setlength{\oddsidemargin}{0in}
%\setlength{\evensidemargin}{0in}
%\setlength{\textwidth}{6.25in}
%\setlength{\topmargin}{-10mm}
%\setlength{\textheight}{9.5in}

\begin{document}

\pagenumbering{roman}


%%%%%%%%  Title page

\thispagestyle{empty}

\mbox{}
\vfill
\begin{center}
{\LARGE\bf OpenRocket technical documentation}

{\large For OpenRocket version 13.05}

\vspace{-3mm}
2013-05-10

\vspace{10mm}
{\Large Sampo Niskanen}


\vspace{80mm}
Based on the Master's thesis \cite{thesis}

\vspace{-1mm}
\mbox{\hspace{-6pt}
\large\it Development of an Open Source model rocket simulation software}

\end{center}

\vfill


\clearpage
\thispagestyle{empty}

\mbox{}
\vfill

\begin{center}
{\Large\bf Thesis or technical documentation?}
\end{center}

The OpenRocket simulation software was originally developed as the
Master's thesis project of Sampo Niskanen, including its written
part
{\it ``Development of an Open Source model rocket simulation software''}
\cite{thesis}.  The thesis is used as the basis of this technical
documentation, which is updated to account for later development in the
software.  This document often still refers to itself as a thesis, as
no systematic updating of this fact has yet been performed.

While the original thesis is available online under a Creative Commons
no-derivatives license, this document is available under a freer
share-alike license.

The latest version of the technical documentation is available on the
OpenRocket website, \url{http://openrocket.sourceforge.net/}.

\vspace{10mm}

\begin{center}
{\Large\bf Version history}
\end{center}

\begin{center}
\begin{tabular}{lp{120mm}}
2010-04-06 & Initial revision.  Updates the roll angle effect on three- and
four-fin configurations in Section~\ref{update-roll-angle}.
(OpenRocket 1.0.0) \\
2011-07-18 & Updated Chapter~\ref{chap-software} for updates in the
software.  (OpenRocket 1.1.6) \\
2013-05-10 & Added Section~\ref{sec-tumbling-bodies} with drag
estimation of tumbling bodies.  (OpenRocket 13.05)
\end{tabular}
\end{center}

\vfill



%%%%%  Quotation


\clearpage
\thispagestyle{empty}

\mbox{}
\vfill
{\it ``No. Coal mining may be your life, but it's not mine. I'm never
  going down there again. I wanna go into space.''}
\vspace{5mm}

\hspace{10mm}\parbox{130mm}{
Amateur rocketeer Homer Hickam, Jr. in the movie October Sky (1999), based
on a true story.\\

Hickam later became an engineer at NASA, working in spacecraft design
and crew training.}
\vfill



%%%%%%%%


\tableofcontents

%\listoffigures
%
%\listoftables

\newpage
\section*{List of symbols and abbreviations}

{\bf Symbols}
\nopagebreak

\begin{tabular}{p{20mm}p{105mm}}

$A$           & Area \\
\Afin         & Area of one fin \\
$A_{\rm plan}$& Planform area \\
\Aref         & Reference area \\
$A_{\rm wet}$ & Wetted area \\
$\AR$         & Aspect ratio of a fin, $2s^2/\Afin$ \\
$c$           & Speed of sound \\
$\bar c$      & Mean aerodynamic chord length of a fin \\
$c(y)$        & Chord length of a fin at spanwise position $y$ \\

$C_A$         & Axial drag force coefficient \\
$C_D$         & Drag force coefficient \\
$C_f$         & Skin friction drag coefficient \\
$C_l$         & Roll moment coefficient \\
$C_{ld}$      & Roll damping moment coefficient \\
$C_{lf}$      & Roll forcing moment coefficient \\

$C_m$         & Pitch moment coefficient \\
\Cma          & Pitch moment coefficient derivative, 
                $\frac{\partial C_m}{\partial \alpha}$ \\

$C_N$         & Normal force coefficient \\
\CNa          & Normal force coefficient derivative,
                $\frac{\partial C_N}{\partial \alpha}$ \\

$d$           & Reference length, the rocket diameter \\
$D$           & Drag force \\
$f_B$         & Rocket fineness ratio, $L/d$ \\
$L$           & The rocket length \\
$m$           & Pitch moment \\
$M$           & Mach number \\
$N$           & Normal force; Number of fins \\
$p$           & Air pressure \\
$r(x)$        & Body or component radius at position $x$ \\
$R$           & Reynolds number \\
$s$           & Spanwise length of one fin \\
$T$           & Air temperature \\
$V$           & Volume \\
$v_0$         & Free-stream velocity \\
$x$, $X$      & Position along the rocket centerline \\
$y$           & Spanwise position \\
\end{tabular}

\begin{tabular}{p{20mm}p{105mm}}
$\alpha$      & Angle of attack \\
$\beta$       & $\sqrt{|M^2-1|}$ \\
$\gamma$      & Specific heat ratio, for air $\gamma=1.4$ \\
$\Gamma_c$    & Fin midchord sweep angle \\
$\delta$      & Fin cant angle \\
$\eta$        & Airflow inclination angle over a fin \\
$\theta$      & Roll angle \\
$\Lambda$     & Dihedral angle between a fin and the direction of airflow \\
$\nu$         & Kinematic viscosity of air \\
$\xi$         & Distance from rotation axis \\
$\rho$        & Density of air \\
$\omega$      & Angular velocity \\

\end{tabular}

\vspace{10mm}
{\bf Abbreviations}
\nopagebreak

\begin{tabular}{p{20mm}p{105mm}}
CFD & Computational fluid dynamics \\
CG  & Center of gravity \\
CP  & Center of pressure \\
LE  & Leading edge \\
MAC & Mean aerodynamic chord \\
RK4 & Runge-Kutta 4 integration method \\
UI  & User interface \\
\end{tabular}




\pagebreak
\pagenumbering{arabic}
\setcounter{page}{1}

\include{chapter-introduction}
\include{chapter-basics-of-model-rockets}
\include{chapter-aerodynamic-properties}
\include{chapter-flight-simulation}
\include{chapter-software}
\include{chapter-experimental}
\include{chapter-conclusion}


\clearpage
\vspace*{1cm}
\section*{Acknowledgments}

I would like to express my deepest gratitude to M.Sc.~Timo Sailaranta
for his invaluable advice and consultation on the aerodynamic
simulation of rockets.  Without his input the creation of the
OpenRocket software and Master's thesis would have been exceedingly
laborious.  I would also like to thank Prof.~Rolf Stenberg for
supervising the writing of the Master's thesis.

I am also deeply grateful for my parents Jouni and Riitta, my entire
family, friends and teachers, who have always encouraged me onwards in
my life and studies.  Above all I would like to thank my brother,
Antti~J. Niskanen, for being an inspiration throughout my life and
also for building the magnetometer logger used in the experimental
flights; and my wife Merli Lahtinen, for her patience and loving
understanding for my passion towards rocketry.




\begin{thebibliography}{99}

\bibitem{thesis} Niskanen, S., {\it Development of an Open Source
  model rocket simulation software}, M.Sc. thesis, Helsinki University
  of Technology, 2009.  Available at
  \url{http://openrocket.sourceforge.net/documentation.html}.

\bibitem{stine} Stine, H., Stine, B., {\it Handbook of Model
  Rocketry}, 7th edition, Wiley, 2004.

\bibitem{barrowman-rd} Barrowman, J., Barrowman, J., The
  theoretical prediction of the center of pressure, {\it National
  Association of Rocketry Annual Meet 8}, 1966.  Available at
\url{http://www.apogeerockets.com/Education/downloads/barrowman_report.pdf},
  retrieved 14.5.2009.

\bibitem{barrowman-thesis} Barrowman, J., {\it The practical
  calculation of the aerodynamic characteristics of slender finned
  vehicles}, M.Sc. thesis, The Catholic University of America, 1967.

\bibitem{rocksim} van Milligan, T., RockSim Model Rocket Design and
  Simulation Software, \url{http://www.apogeerockets.com/RockSim.asp},
  retrieved 14.5.2009.

\bibitem{oss-principles} Coar, K., The Open Source Definition
  (Annotated), \url{http://www.opensource.org/docs/definition.php},
  retrieved 14.5.2009.

\bibitem{openrocket} Niskanen, S., The OpenRocket web-site,
  \url{http://openrocket.sourceforge.net/}, retrieved 25.5.2009.

\bibitem{nar-safety-code} Anon., Model Rocket Safety Code,
  \url{http://www.nar.org/NARmrsc.html}, retrieved 14.5.2009.

\bibitem{all-certified-motors} Anon., Combined CAR/NAR/TRA Certified
  Rocket Motors List,
  \url{http://www.nar.org/SandT/pdf/CombinedList.pdf}, retrieved 14.5.2009.

\bibitem{thrust-curve-database} Coker, J., ThrustCurve Hobby Rocket
  Motor Data, \url{http://www.thrustcurve.org/}, retrieved 14.5.2009.

\bibitem{D12-curve} Kane, J., Estes D12,
  \url{http://www.nar.org/SandT/pdf/Estes/D12.pdf}, retrieved 14.5.2009.

\bibitem{haisunaata-launch} Puhakka, A., Haisunäätä---suomalainen
  hybridirakettiprojekti (in Finnish),
  \url{http://haisunaata.avaruuteen.fi/}, retrieved 14.5.2009.

\bibitem{galejs} Galejs, R., Wind instability---What Barrowman left
  out,
  \url{http://projetosulfos.if.sc.usp.br/artigos/sentinel39-galejs.pdf},
  retrieved 14.5.2009.

\bibitem{advanced-model-rocketry} Mandell, G., Caporaso, G., Bengen,
  W., {\it Topics in Advanced Model Rocketry}, MIT Press, 1973.

\bibitem{hoerner} Hoerner, S., {\it Fluid-dynamic drag}, published by
  the author, 1965.
% FLUID-DYNAMIC DRAG
% Practical Information on AERODYNAMICDRAG and HYDRODYNAMIC RESISTANCE
% Sighard F. Hoerner (Dr.-Ing.)
% Published by the Author 1958
%
% Chap II - Skin-friction drag
%   laminaarinen, turbulentti, ym.
% Chap III - Pressure drag
%   forebudy pressure drag for different shapes
%   Base drag C_DB = 0.029/sqrt(C_fB)  forebody-drag coefficient C_fB
% Chap V - Drag of surface imperfections
%   Drag due to surface roughtness
%   Critical roughness
%   Page 5-8, Drag of Individual Protuberances
%     neliskanttinen pala, pituus < korkeus -> CD=1.20
%                          pituus > 2*kork  -> CD=0.74
%                     suhteutettu etupinta-alaan
%     From ref. Tillmann, Rpt KW Inst. Göttingen, Dec 1944
% Chap VII - Drag due to lift
% Chap VIII - Interference drag
%   Pairs of bodies
% Chap X - Hydrodynamic drag
%   sivu 10-3, siivekkeiden profiilimuotoja!!!
% Chap XIII - Drag of airplane components and accessories
%   Drag of external loads
%   Parachutes
% Chap XV-XVII - subsonic, transsonic, supersonic

\bibitem{barrowman-elliptical-fins} Barrowman, J., Elliptical Fin
  C.P. Equations, {\it Model Rocketry} (Nov 1970).  Available at
  \url{http://www.argoshpr.ch/articles/pdf/EllipticalCP.jpg},
  retrieved 14.5.2009.

\bibitem{appl-comp-aero-fins} Mason, W., Applied Computational
  Aerodynamics,
  \url{http://www.aoe.vt.edu/~mason/Mason_f/CAtxtTop.html}, 
  {\bf pp. A-27--A-28}, retrieved 14.5.2009.

\bibitem{fleeman} Fleeman, E., {\it Tactical missile design}, 2nd
  edition, p.~33, AIAA, 2006.

\bibitem{diederich} Diederich, F., {\it A plan-form parameter for
 correlating certain aerodynamic characteristics of swept ings},
 NACA-TN-2335, 1951.

\bibitem{barrowman-fin} Barrowman, J., {\it FIN A computer program for
  calculating the aerodynamic characteristics of fins at supersonic
  speeds}, {\it NASA-TM X-55523}, 1966.

\bibitem{pettis} Pettis, W., {\it Aerodynamic Characteristics of
  Multiple Fins of Rectangular Planform on a Body of Revolution at
  Mach Numbers of 1.48 to 2.22}, RD-TM-67-5, US Army Missile
  Command, 1967.

\bibitem{experimental-transonic} Ferris, J., {\it Static stability
  investigation of a single- stage sounding rocket at Mach numbers
  from 0.60 to 1.20}, NASA-TN-D-4013, 1967.

\bibitem{triform-fin-data} Monta, W., {\it Aerodynamic
  characteristics at mach numbers from 1.60 to 2.16 of a blunt-nose
  missile model having a triangular cross section and fixed triform
  fins}, NASA-TM-X-2340, 1971.

\bibitem{MIL-HDBK} Anon., {\it Design of aerodynamically stabilized
  free rockets}, MIL-HDBK-762, US Army Missile Command, 1990.

\bibitem{handbook-supersonic-aerodynamics} Anon., {\it Handbook of
  supersonic aerodynamics, Section 8, Bodies of revolution}, 
  NAVWEPS REPORT 1488, 1961.

\bibitem{second-order-shock-expansion-method} Syverston, C., Dennis,
  D., {\it A second-order shock-expansion method applicable to bodies
  of revolution near zero lift}, NACA-TR-1328, 1957.

\bibitem{international-standard-atmosphere} Anon., {\it Standard
  Atmosphere}, ISO~2533:1975, International Organization for
  Standardization, 1975.

\bibitem{US-standard-atmosphere} Anon., {\it U.S. Standard Atmosphere
  1976}, NASA-TM-X-74335; NOAA-S/T-76-1562, 1976.

\bibitem{wiki-ISA-layers}  Anon., International Standard Atmosphere,
  \url{http://en.wikipedia.org/wiki/International_Standard_Atmosphere},
  retrieved 14.5.2009.

\bibitem{wind-energy-handbook} Burton, T., Sharpe, D., Jenkins, N.,
  Bossanyi, E., {\it Wind Energy Handbook}, Wiley, 2001.

\bibitem{pink-filter}  Kasdin, J., Discrete Simulation of Colored
  Noise and Stochastic Processes and $1/f^\alpha$ Power Law Noise
  Generation, {\it Proceedings of the IEEE}, {\bf 83}, No.~5 (1995),
  p. 822.

\bibitem{wiki-euler-angles} Anon., Euler angles,
  \url{http://en.wikipedia.org/wiki/Euler_angles}, retrieved 14.5.2009.

\bibitem{wiki-euler-rotation-theorem} Anon., Euler's rotation theorem,
  \url{http://en.wikipedia.org/wiki/Euler's_rotation_theorem},
  retrieved 14.5.2009.

\bibitem{wiki-quaternion-rotations} Anon., Quaternions and spatial
  rotation,
  \url{http://en.wikipedia.org/wiki/Quaternions_and_spatial_rotation},
  retrieved 14.5.2009.

\bibitem{wiki-moments-of-inertia} Anon., List of moments of inertia,
  \url{http://en.wikipedia.org/wiki/List_of_moments_of_inertia},
  retrieved 14.5.2009.

\bibitem{pollux-wind-tunnel} Niskanen, S., Polluxin tuulitunneli (in
  Finnish), \url{http://pollux.tky.fi/tuulitunneli.html}, retrieved 14.5.2009.

\bibitem{gnu-gpl} Anon., GNU General Public License, Version 3,
  \url{http://www.gnu.org/copyleft/gpl.html}, retrieved 14.5.2009.

\bibitem{java-packages} Anon., Java Language Specification, Chaper 7,
  Packages,
 \url{http://java.sun.com/docs/books/jls/third_edition/html/packages.html#7.7},
  retrieved 14.5.2009.

\bibitem{GZIP} Deutsch, P., {\it GZIP file format specification
  version 4.3}, RFC~1952, \url{http://www.ietf.org/rfc/rfc1952.txt},
  1996.  Retrieved on 14.5.2009.

\bibitem{perfectflite} Anon., Affordable instrumentation for (sm)all
  rockets, \url{http://www.perfectflite.com/}, retrieved 14.5.2009.

\bibitem{weco-feuerwerk} Anon., WECO Feuerwerk,
  \url{http://www.weco-feuerwerk.de/}, retrieved 14.5.2009.

\bibitem{estes} Anon., Estesrockets.com,
  \url{http://www.estesrockets.com/}, retrieved 14.5.2009.

\bibitem{sf-thrustcurves} Anon., Schubdiagramme SF,
  \url{http://www.raketenmodellbautechnik.de/produkte/Motoren/SF-Motoren.pdf},
  14.5.2009.

\bibitem{experimental-supersonic} Babb, C., Fuller, D., {\it Static
  stability investigation of a sounding-rocket vehicle at Mach numbers
  from 1.50 to 4.63}, NASA-TN-D-4014, 1967.

\bibitem{nosecone-cd-data} Stoney, W., {\it Collection of Zero-Lift
  Drag Data on Bodies of Revolution from Free-Flight Investigations}, 
  NASA-TR-R-100, 1961.

\bibitem{streamer-optimization} Kidwell, C., Streamer Duration
  Optimization: Material and Length-to-Width Ratio, {\it National
  Association of Rocketry Annual Meet 43}, 2001.  Available at
  \url{http://www.narhams.org/library/rnd/StreamerDuration.pdf},
  retrieved 14.5.2009.

\end{thebibliography}




\appendix

\include{chapter-appendices}




\end{document}
